---
title: Practical 3
subtitle: Functions & Formulas
author: Mykola Chuprynskyy
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 2
  
---

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
```


```{r, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

***

# 1 Write your own `my_cumsum()` function

The base R function cumsum() returns the cumulative sum of its argument, e.g.


```{r}
cumsum(1:4)
```


a. Write the function `my_cumsum()` that does the same.
```{r}
my_cumsum <- function(x) {
  
    sum <- cumsum(x)
    
  return(sum)
  }

```

b. Show that the function works for the argument `-1:4`
```{r}
my_cumsum(-1:4)
```

# 2 Bootstrap

a. Source the `bootstrap()` function of the lecture slides.
```{r}
bootstrap <- function(formula, data, B = 1000) {  
  
  pars   <- lm(formula = formula, data = data) %>% 
    coef() %>% 
    names()
  
  output <- matrix(nrow = B, ncol = length(pars), 
                     dimnames = list(NULL, pars))
  
  for (i in 1:B) {
    
    bootsample  <- slice_sample(.data = data, n = nrow(data), replace = TRUE)
    
    bootfit     <- lm(formula = formula, data = bootsample)
    
    output[i, ] <- bootfit$coef
  }
  
  return(output)
}
```

b. Fit the linear model `mpg ~ disp + drat + qsec` of the `mtcars` data and save the `lm` object as `fit`.
```{r}
fit <- lm(mpg ~ disp + drat + qsec, data = mtcars)
bootstrap_results <- bootstrap(mpg ~ disp + drat + qsec, data = mtcars, B = 1000)
```

c. Display the summary of this model.
```{r}
summary(fit)

```

d. Save the bootstrap object for the same model.
```{r}
save(bootstrap_results, file = "bootstrap_results.RData")
```

e. Make an object with the bootstrap confidence intervals and one with the theoretical confidence intervals, and compare both.
```{r}
# Calculate the 95% bootstrap confidence intervals
bootstrap_ci <- apply(bootstrap_results, 2, function(coefs) {
  quantile(coefs, probs = c(0.025, 0.975))
})

# Convert to a more readable format
bootstrap_ci_df <- as.data.frame(t(bootstrap_ci))
colnames(bootstrap_ci_df) <- c("Lower", "Upper")
print("Bootstrap Confidence Intervals:")
print(bootstrap_ci_df)

# Calculate the 95% theoretical confidence intervals
theoretical_ci <- confint(fit, level = 0.95)

print("Theoretical Confidence Intervals:")
print(theoretical_ci)

```

f. In a 2 x 2 plot array, plot the bootstrap sampling distributions of the four parameters in red, and the theoretical sampling distribution of the parameters in blue. For the latter, use `stat_function(fun = dnorm, args = list(mean = ..., sd = ...))`, using the mean and sd values of the corresponding `lm` model estimates.
```{r}

num_params <- ncol(bootstrap_ci_df)


plots <- lapply(1:4, function(i) {
 
  param_index <- ((i - 1) %% num_params) + 1
  param_mean <- mean(bootstrap_ci_df[[param_index]])
  param_sd <- sd(bootstrap_ci_df[[param_index]])
  
  ggplot() +
    geom_density(data = data.frame(Value = bootstrap_ci_df[[param_index]]), aes(x = Value), fill = "red", alpha = 0.5, color = "red") +
    stat_function(fun = dnorm, args = list(mean = param_mean, sd = param_sd), color = "blue") +
    labs(title = colnames(bootstrap_ci_df)[param_index]) +
    theme_minimal()
})

grid.arrange(grobs = plots, ncol = 2)
```

***

End of practical