---
title: "HW1"
author: "Mykola Chuprinskiy"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# This will plot the item frequencies and print out the top 3 most frequent items sold.
## Load necessary libraries

```{r}
library(arules)
library(arulesViz)
```

## Import the data
```{r}
DAT <- read.table('Online Retail BIM.dat', header = TRUE)
```

## Convert the data to a matrix, excluding the invoice numbers
```{r}
BIM <- as.matrix(DAT[-1])
```

## Convert the matrix to a 'transactions' class
```{r}
BIM <- as(BIM, "transactions")
```

## Plot relative item frequency (support)
```{r}
itemFrequencyPlot(BIM, topN = 20, type = "relative", main = "Relative Item Frequency Plot")
```

## To determine the top 3 items with the highest support
```{r}
item_support <- itemFrequency(BIM, type = "absolute")
top_3_items <- sort(item_support, decreasing = TRUE)[1:3]
top_3_items
```


# Determine which two association rules containing only two items have the highest support.
```{r}
# Apply the apriori algorithm
rules <- apriori(BIM, parameter = list(supp = 0.01, conf = 0.5, minlen = 2, maxlen = 2))

# Sort the rules by support and inspect the top 2
sorted_rules <- sort(rules, by = "support", decreasing = TRUE)
inspect(sorted_rules[1:2])

```

# How many items does the association rule with the highest confidence contain?
```{r}
rules_by_conf <- sort(rules, by = "confidence", decreasing = TRUE)

inspect(rules_by_conf[1])

length(rules_by_conf[1]@lhs) + length(rules_by_conf[1]@rhs)

```

# Which items are important antecedents of item 4050? What does this mean?
```{r}
rules_4050 <- apriori(BIM, parameter = list(supp = 0.01, conf = 0.5), 
                      appearance = list(lhs = 'i4050', default = "rhs"))
inspect(rules_4050)

sorted_rules_4050 <- sort(rules_4050, by = "confidence", decreasing = TRUE)
sorted_rules_4050[1:2]

```

# Which items are important consequents of item 1655? What does this mean?
```{r}
rules_1655 <- apriori(BIM, parameter = list(supp = 0.01, conf = 0.5),
                      appearance = list(lhs = "i1655", default = "rhs"))
inspect(rules_1655)

sorted_rules_1655 <- sort(rules_1655, by = "confidence", decreasing = TRUE)
inspect(sorted_rules_1655[1:3])

```

# Are there two items that are substitutes for each other? Why?
```{r}
substitute_rules <- subset(rules, lift < 1.1)
inspect(substitute_rules)

```

# Make a scatter plot and a two-key plot for association rules with minimum support 0.01 and minimum confidence 0.5.
```{r}
rules_viz <- apriori(BIM, parameter = list(supp = 0.01, conf = 0.5))

plot(rules_viz, method = "scatterplot", measure = c("support", "confidence"), shading = "lift")

plot(rules_viz, method = "two-key plot")

```


